Neighborhoods and Felony Disenfranchisement: The Case of New York City

### 2019-07-24 markdown/part2_standalone.Rmd
write

Below, I employ a logistic regression to investigate the impact of Executive Order 181 on turnout among all parolees discharged after the policy went into effect. Because not all individuals had their rights restored, I follow this analysis with an instrumental-variables approach to determine the effect of rights restoration on individuals whose rights were in fact formally restored prior to discharge from parole.

@@

Although it perhaps seems counterintuitive to throw out many of our observations, ignoring all individuals last discharged from parole prior to 2017 give us an important asset: it allows us to conceptualize Executive Order 181 as a natural experiment. Whether an individual was discharged from parole before or after May 21^st^, 2018, is akin to a randomly assigned treatment. Any observed difference in turnout between individuals discharged before and after late May can therefore be attributed to the Executive Order.

In Table \ref{tab:to-18-logit}, I present the results of an individual-level logistic regression exploring whether individuals who were discharged on or after May 21^st^, 2018, turned out at higher rates than those discharged earlier. The models include all individuals discharged from parole between January 1^st^, 2017, through October 10^th^, 2018 (the registration deadline in New York State).

@@

Model 1 in Table \ref{tab:to-18-logit} formalizes the trend presented in Figure \ref{fig:to-18-recent-chart} by controlling only for whether an individual was discharged after Executive Order 181 went into effect. Model 2 also controls for individual-level characteristics: sex, age on November 6^th^, 2018, county of incarceration, and race. Model 3 adds sentence-specific information to Model 2: the number of counts in the individual's most recent sentence, the amount of time they spent on parole, and the class(es) of felony for which they were convicted. Table \ref{tab:to-18-logit} makes clear that formerly incarcerated men were far less likely to vote than formerly incarcerated women; that older formerly incarcerated individuals were more likely to cast a ballot; and individuals who spent longer on parole were more likely to participate in the midterm election.

@@

Table \ref{tab:to-18-logit} indicates that Executive Order was successful at increasing turnout among all formerly disenfranchised individuals. Although this is important information for policymakers and advocates hoping to increase the political representation of formerly incarcerated individuals as a whole, it does not shed light on the extent to which having one's voting rights restored prior to discharge from parole impacts one's propensity to vote. To answer that question, we must specifically control for whether an individual did have his rights restored before he was discharged from parole.

>>>>

Below, I employ a logistic regression to investigate the impact of Executive Order 181 on turnout among all parolees discharged after the policy went into effect. However, not all individuals discharged after the Executive Order went into effect had their rights restored. Because an individual's likelihood of having his rights restored prior to discharge from parole may be correlated with his propensity to vote, I also employ an instrumental-variables approach to estimate the effect of the Executive Order on turnout.

@@

Although it perhaps seems counterintuitive to throw out many of our observations, ignoring all individuals last discharged from parole prior to 2017 give us an important asset: it allows us to conceptualize Executive Order 181 as a natural experiment. Whether an individual was discharged from parole before or after May 21^st^, 2018, is akin to a randomly assigned treatment. There is no evidence that individuals discharged from parole after this date would have turned out at different rates in the absence of the Executive Order. Any observed difference in turnout between individuals discharged before and after late May can therefore be attributed to the Executive Order.

Table \ref{tab:to-18-logit} presents models estimating individual-level turnout. The dependent variable is binary --- it equals 1 if an individual cast a ballot, and 0 if they did not. Because of the binary nature of the dependent variable, I fit the models using a logistic regression. The models include all individuals discharged from parole between January 1^st^, 2017, through October 10^th^, 2018 (the registration deadline in New York State).

@@

Model 1 in Table \ref{tab:to-18-logit} formalizes the trend presented in Figure \ref{fig:to-18-recent-chart} by controlling only for whether an individual was discharged after Executive Order 181 went into effect. Model 2 adds individual-level demographic characteristics: sex, age on November 6^th^, 2018, county of incarceration, and race. Model 3 adds sentence-specific information to Model 2: the amount of time an individual spent on parole and the class(es) of felony for which they were convicted. Table \ref{tab:to-18-logit} makes clear that formerly incarcerated men were far less likely to vote than formerly incarcerated women; that older formerly incarcerated individuals were more likely to cast a ballot; and individuals who spent longer on parole were more likely to participate in the midterm election.

@@

Table \ref{tab:to-18-logit} indicates that Executive Order was successful at increasing turnout among all formerly disenfranchised individuals. Although this is important information for policymakers and advocates hoping to increase the political representation of formerly incarcerated individuals as a class, it does not shed light on the extent to which having one's voting rights restored prior to discharge from parole impacts one's propensity to vote. To answer that question, we must look specificially at whether an individual did have his rights restored before he was discharged from parole.
